One page description

### 2016-12-03 publications/Journal/3-features.tex
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PDEVS with \pSim * changed mentions of "sequential" and "parallel" to single core and multi core where applicable. - - -> might need review to check if I didn't miss anything

These different protocols relate to three different model characteristics.
Conservative synchronization for when high lookahead exists between different nodes, and barely any blocking is necessary.
Optimistic synchronization for when lookahead is unpredictable, or there are rare (almost) instantaneous events.
Finally, sequential simulation is still required for models where parallelism is bad. 
It is possible for the synchronization overhead to become larger than the achieved parallelism, resulting in slower simulation than sequential.

@@

% protocol or inherent parallelism?
\subsubsection{pSim}
In our tool each simulation core, regardless of inter core synchronization, is capable of executing concurrent transitions in parallel. A configurable number of threads is allocated to optimally divide the processing load of concurrent transitions. The parallel execution of transitions introduces some overhead in thread pooling and locking, and disallows some optimizations in optimally rescheduling models. The concurrency of events and the combined computational load of the transitions has to outweigh this overhead to obtain a significant speedup, as we will demonstrate in Section~\textsc{\nameref{sec:4-performance}}.

\subsection{Synchronization Protocol Transparency}
We define synchronization protocol transparency as having a single model, which can be executed on each supported synchronization kernel, without any modifications.
User should thus only provide one model, implemented in C++11, which can be simulated using sequential execution, conservative synchronization, or optimistic synchronization.
The synchronization protocol to use is a simple configuration option.
The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.

@@

\subsubsection{Memory Management}
\label{sec:4-subsec:overhead-pgraph:memory}
Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}, since memory bandwidth doesn't increase as fast as the number of cores using it.
While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
A model written for sequential simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.

Furthermore, allocating and deallocating memory by making calls to the operating system, as is typically done by calls such as \texttt{malloc}, happens sequentially.
To counter this, our memory allocators are backed by a thread-aware pooling library.
In a sequential simulation kernel no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.

@@

\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in parallel simulation kernels and arena-style allocators for sequential simulation.
The latter can be faster, but at the cost of additional configuration.
The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, reducing lock contention in \texttt{malloc} calls.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
    %TODO: Add value on top of bars
    \caption{Effect of memory allocators on parallel execution time, using optimistic synchronization.}
    \label{fig:memallocators_parallel}
\end{figure}

@@

\subsubsection{Random Number Generators}
Random Number Generators (RNG) are another aspect of the program that can limit parallelization.
All accesses to the RNG will result in the modification of a global (\textit{i.e.}, shared between threads) variable.
This easily becomes a bottleneck in simulation, since random numbers are a common occurence in simulation~\cite{Random}.
As such, a nontrivial amount of time in a simulation is often spent waiting for an RNG.

@@

We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default Standard Template Library (STL) random number generator with optimistic synchronization.
This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
\textit{Dxex}'s sequential and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.

Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in sequential simulation is three times faster using any of the three alternatives, only the STL RNG is slow.
For parallel simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between sequential and parallel simulation.
Conservative synchronization is almost unaffected by changing the RNG.

>>>>

These different protocols relate to three different model characteristics.
Conservative synchronization for when high lookahead exists between different nodes, and barely any blocking is necessary.
Optimistic synchronization for when lookahead is unpredictable, or there are rare (almost) instantaneous events.
Finally, single core simulation is still required for models where parallelism is bad. 
It is possible for the synchronization overhead to become larger than the achieved parallelism, resulting in slower simulation than single core.

@@

% protocol or inherent parallelism?
\subsubsection{\pSim}
In our tool each simulation core, regardless of inter core synchronization, is capable of executing concurrent transitions in parallel. A configurable number of threads is allocated to optimally divide the processing load of concurrent transitions. The parallel execution of transitions introduces some overhead in thread pooling and locking, and disallows some optimizations in optimally rescheduling models. The concurrency of events and the combined computational load of the transitions has to outweigh this overhead to obtain a significant speedup, as we will demonstrate in Section~\textsc{\nameref{sec:4-performance}}.

\subsection{Synchronization Protocol Transparency}
We define synchronization protocol transparency as having a single model, which can be executed on each supported synchronization kernel, without any modifications.
User should thus only provide one model, implemented in C++11, which can be simulated using single core execution, conservative synchronization, or optimistic synchronization.
The synchronization protocol to use is a simple configuration option.
The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.

@@

\subsubsection{Memory Management}
\label{sec:4-subsec:overhead-pgraph:memory}
Memory management is traditionally seen as one of the major bottlenecks in parallel computation~\cite{Memory}, since memory bandwidth doesn't increase as fast as the number of cores using it.
While this is always a problem, it is aggravated in \textit{dxex} by providing automatic memory management for events and states.
A model written for single core simulation will run correctly in a conservative or optimistic simulation without altering (from the point of view of the modeller) the (de)allocation semantics of events or states.

Furthermore, allocating and deallocating memory by making calls to the operating system, as is typically done by calls such as \texttt{malloc}, happens sequentially.
To counter this, our memory allocators are backed by a thread-aware pooling library.
In a single core simulation kernel no allocated event persists beyond a single time advance, even allowing the use of an arena-style allocator.
Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime.

@@

\textit{Dxex} uses \textit{Boost Pool}~\cite{boostpool} allocators in multi core simulation kernels and arena-style allocators for single core simulation.
The latter can be faster, but at the cost of additional configuration.
The allocators are supplemented by the library \textit{tcmalloc}~\cite{tcmalloc}, reducing lock contention in \texttt{malloc} calls.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/memory_allocators_parallel.eps}
    %TODO: Add value on top of bars
    \caption{Effect of memory allocators on multi core execution time, using optimistic synchronization.}
    \label{fig:memallocators_parallel}
\end{figure}

@@

\subsubsection{Random Number Generators}
Random Number Generators (RNG) are another aspect of the program that can limit parallelization.
All accesses to the RNG will result in the modification of a global (\textit{i.e.}, shared between threads) variable.
This easily becomes a bottleneck in simulation, since random numbers are a common occurrence in simulation~\cite{Random}.
As such, a nontrivial amount of time in a simulation is often spent waiting for an RNG.

@@

We see in Figure~\ref{fig:Queuerngspeedup} that storing the RNG in the state is very expensive for the default Standard Template Library (STL) random number generator with optimistic synchronization.
This is primarily caused by the significant difference in size: 2504 bytes for the STL random number generator, and 24 bytes for the Tina random number generator.
\textit{Dxex}'s single core and conservative kernels are insensitive to storing the RNG object in the atomic model state, since no copying/state saving occurs in \textit{dxex} conservative simulation.
The optimistic kernel is clearly affected, as it needs to copy more bytes in every transition due to state saving.

Figure~\ref{fig:Queuerngspeedup} shows that \textit{dxex} in single core simulation is three times faster using any of the three alternatives, only the STL RNG is slow.
For multi core simulation, the synchronization overhead seems to be the main bottleneck, as seen by the big speedup gap between single core and multi core simulation.
Conservative synchronization is almost unaffected by changing the RNG.

### 2016-12-03 publications/Journal/4-performance.tex
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PDEVS with \pSim * changed mentions of "sequential" and "parallel" to single core and multi core where applicable. - - -> might need review to check if I didn't miss anything

In this section, we evaluate the performance of different synchronization protocols in \textit{dxex}.
We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVStoneJournal}, to show that our modularity does not impede performance.
CPU time and memory usage is compared for both sequential and parallel simulation.

We start off with a comparison of sequential simulation, to show how \textit{adevs} and \textit{dxex} relate in this simple case.
For the parallel simulation benchmarks, results are presented for both conservative and optimistic synchronization.

@@

\subsection{Sequential Simulation}
We start by evaluating sequential simulation performance, in order to obtain a baseline for our comparison of parallel simulation performance.

@@

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/queue_fixed_sequential.eps}
	\caption{Queue benchmark results for sequential simulation.}
	\label{fig:Queue_benchmark_seq}
\end{figure}

@@

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/interconnect_sequential.eps}
	\caption{Interconnect benchmark results for sequential simulation.}
	\label{fig:Interconnect_benchmark}
\end{figure}

@@

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/phold_sequential.eps}
	\caption{PHold benchmark results for sequential simulation.}
	\label{fig:Phold_benchmark}
\end{figure}

\subsection{Parallel Simulation}
We now continue by describing our parallel simulation performance for different synchronization protocols, and compared to \textit{adevs}.
The speedup of \textit{adevs} is computed with the corresponding \textit{dxex} sequential benchmark.
This was done to take into account the performance difference observed in sequential simulation.
As such, the highest speedup indicates the fastest results among all tools, independent of sequential simulation results.

\subsubsection{Queue}
The Queue model is one single chain of models, resembling a pipeline.
This structure can be exploited to prevent cyclic dependencies in the parallel simulation.

Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to sequential simulation for a fixed problem size (\textit{i.e.}, strong scaling).
As the number of kernels increases, the optimistic protocol quickly becomes the worst choice.
This is mainly caused by the pipeline structure of the model: the last models in the queue only respond to incoming messages and therefore have to be rolled back frequently.
The difference between \textit{dxex} conservative protocol and \textit{adevs} becomes smaller when more and more cores are used.
The same effect can be seen when the problem size is increased in tandem with the number of used cores (\textit{i.e.}, weak scaling) in Figure~\ref{fig:Queue_plot_weak}.

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
	\caption{Queue model strong scaling speedup compared to \textit{dxex} sequential.}
	\label{fig:Queue_plot_strong}
\end{figure}

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/queue_fixed_weak_speedup.eps}
	\caption{Queue model weak scaling speedup compared to \textit{dxex} sequential.}
	\label{fig:Queue_plot_weak}
\end{figure}

\subsubsection{Interconnect}
\label{subsec:parallelinterconnect}
In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
The number of models is now kept constant at eight.
Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and the increasing number of rollbacks in optimistic simulation.
All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in parallel.

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
    %TODO cyclic allocation in title?
	\caption{Interconnect benchmark results for parallel simulation.}
	\label{fig:interconnect_benchmark_parallel}
\end{figure}

@@

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
    \caption{PHOLD benchmark results for parallel simulation using four kernels, with varying percentage of remote events.}
\end{figure}

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/phold_priority.eps}
	\caption{PHOLD benchmark results for parallel simulation using four kernels, with varying amount of high-priority events.}
	\label{fig:phold_priority}
\end{figure}

@@

\subsection{Comparison to the abstract simulator}
%reference pSim
The abstract simulator of \textsf{Parallel DEVS} included its own notion of parallelism: transition functions that happen at the same point in simulated time can be trivially parallelized. In this work we refer to this approach as pSim.
This parallelization is trivial, in the sense that there is no need for synchronization protocols.
The parallelization of the transition functions (from now on called the ``inherent parallelism'') is, at least in \textit{dxex}, independent of the use of a synchronization protocol.
That is, both can be used simultaneously.
Following the theoretical analysis published in~\cite{amdahlpdevs}, a comparison between this inherent parallelism and the use of synchronization protocols is warranted.

\paragraph{Model}
We have opted to use the \textit{Queue} model for this comparison, as it allows for many interesting configurations.
%TODO what is meant with this allocation of threads and synchronized kernels?
% 4 Kernels running on a dedicated thread, with per kernel 4 threads allocated to use for PDEVS.
The effect of combining 4 synchronized kernels, with each 4 threads for PDEVS, is contrasted with 16 threads for PDEVS and 16 kernels for conservative and optimistic.
The first scenario runs each kernel on a dedicated thread synchronized with the other 3 kernels. Each kernel can use up to 4 threads for PDEVS. In the second scenario PDEVS can use up to 16 threads while running in a sequential kernel. Finally 16 synchronized kernels are used with one thread per kernel.
This allows us to observe which is more efficient in obtaining a speedup.
We simulate a computational load by an optional call to sleep in the transition functions set at 5 milliseconds.
The model is configured with depth 4 and width 300 if the transition function has no load, and width 50 if the computational load is active.
In our configuration, an imminent model generates output, resulting in the receiving model becoming imminent.
The probability that an internal transition in a model generates output and is connected to a receiving model here is 1. With respect to the theoretical analysis ~\cite{amdahlpdevs} this gives a parameter value $q = 1$. 
The probability that a model becomes imminent (i.e. is scheduled to execute an internal transition at a given point in time) in this model is equal to 1 when a fixed timeadvance is used,  %todo p calculation
$\frac{1}{timestep}$ when a random time advance is used, this corresponds with $p = 1$ and $p = \frac{1}{100}$ respectively.

The benchmarks are run sufficiently long enough to guarantee that the frequency of internal and external transitions is equal between random and non-random benchmarks. In our model each model will be executing internal and external transitions, creating an ideal use case to evaluate the speedup inherent parallelism can obtain.
The key difference is that although the event frequency is the same, their concurrency is not. 
Models are equally distributed over the kernels and between threads in PDEVS.
The coefficient of variation of our results is less than 1\%.
The communication overhead is hard to estimate, but given our coefficient of variation, we can expect this overhead to be constant.

@@

\paragraph{Concurrent events}
First we create a model where all transitions happen simultaneously, with a significant computational load in the transition functions.
In Figure~\ref{fig:pdevs_plot_fixed_sleep}, we observe that exploiting only the inherent parallelism, results in a speedup of about 10.  
Simulation with conservative and optimistic synchronization is run first with 16 kernels, then with 4 kernels having 4 PDEVS threads each. 
Both synchronization protocols achieve a speedup similar to only exploiting the inherent parallelism.
Combining the protocols with the inherent parallelism gives near identical results.
In this scenario there is no real advantage between the different parallel configurations.

@@

\paragraph{Computational load}
Finally, we remove the computation load from the transition function, in combination with many concurrent events.
Figure~\ref{fig:pdevs_plot_no_sleep} shows that in sequential simulation, the overhead of thread management and shared memory communication is crippling for performance.
Even though many events occur concurrently, the actual computation being done in the transition function does not outweigh the overhead, resulting in much slower simulation than sequential execution of transition functions.
When combined with synchronization protocols, performance is still reasonable, though the use of the inherent parallelism is not beneficial either way.
Both synchronization protocols achieve a reasonable speedup without PDEVS, as the overhead of thread management (creation and destruction) is avoided.
These results are similar to~\cite{Himmelspach}.

@@

Conservative simulation always uses more memory than sequential simulation, as is to be expected.
Additional memory is required for the multiple threads, but also to store all events that are processed simultaneously.

@@

Second, the allocation of models is nontrivial and has a significant impact on performance.
While our parallel speedup for the Queue model, for example, was rather high, this is mostly due to characteristics of the model: the dependency graph does not contain any cycles.
When cycles were introduced, as in the Interconnect model, performance became disastrous.

>>>>

In this section, we evaluate the performance of different synchronization protocols in \textit{dxex}.
We also compare to \textit{adevs}~\cite{adevs}, currently one of the most efficient simulation kernels~\cite{DEVStoneJournal}, to show that our modularity does not impede performance.
CPU time and memory usage is compared for both single core and multi core simulation.

We start off with a comparison of single core simulation, to show how \textit{adevs} and \textit{dxex} relate in this simple case.
For the multi core simulation benchmarks, results are presented for both conservative and optimistic synchronization.

@@

\subsection{Signle Core Simulation}
We start by evaluating single core simulation performance, in order to obtain a baseline for our comparison of multi core simulation performance.

@@

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/queue_fixed_sequential.eps}
	\caption{Queue benchmark results for single core simulation.}
	\label{fig:Queue_benchmark_seq}
\end{figure}

@@

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/interconnect_sequential.eps}
	\caption{Interconnect benchmark results for single core simulation.}
	\label{fig:Interconnect_benchmark}
\end{figure}

@@

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/phold_sequential.eps}
	\caption{PHold benchmark results for single core simulation.}
	\label{fig:Phold_benchmark}
\end{figure}

\subsection{Parallel Simulation}
We now continue by describing our multi core simulation performance for different synchronization protocols, and compared to \textit{adevs}.
The speedup of \textit{adevs} is computed with the corresponding \textit{dxex} single core benchmark.
This was done to take into account the performance difference observed in single core simulation.
As such, the highest speedup indicates the fastest results among all tools, independent of single core simulation results.

\subsubsection{Queue}
The Queue model is one single chain of models, resembling a pipeline.
This structure can be exploited to prevent cyclic dependencies in the multi core simulation.

Figure~\ref{fig:Queue_plot_strong} shows the speedup compared to single core simulation for a fixed problem size (\textit{i.e.}, strong scaling).
As the number of kernels increases, the optimistic protocol quickly becomes the worst choice.
This is mainly caused by the pipeline structure of the model: the last models in the queue only respond to incoming messages and therefore have to be rolled back frequently.
The difference between \textit{dxex} conservative protocol and \textit{adevs} becomes smaller when more and more cores are used.
The same effect can be seen when the problem size is increased in tandem with the number of used cores (\textit{i.e.}, weak scaling) in Figure~\ref{fig:Queue_plot_weak}.

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/queue_fixed_strong_speedup.eps}
	\caption{Queue model strong scaling speedup compared to \textit{dxex} single core.}
	\label{fig:Queue_plot_strong}
\end{figure}

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/queue_fixed_weak_speedup.eps}
	\caption{Queue model weak scaling speedup compared to \textit{dxex} single core.}
	\label{fig:Queue_plot_weak}
\end{figure}

\subsubsection{Interconnect}
\label{subsec:parallelinterconnect}
In the Interconnect model, we determine how broadcast communication is supported across multiple nodes.
The number of models is now kept constant at eight.
Results are shown in Figure~\ref{fig:interconnect_benchmark_parallel}.
When the number of nodes increases, performance decreases due to increasing contention in conservative simulation and the increasing number of rollbacks in optimistic simulation.
All models depend on each other and have no computational load whatsoever, negating any possible performance gain by executing the simulation in a multi core setting.

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/interconnect_parallel.eps}
    %TODO cyclic allocation in title?
	\caption{Interconnect benchmark results for multi core simulation.}
	\label{fig:interconnect_benchmark_parallel}
\end{figure}

@@

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/phold_remotes.eps}
    \caption{PHOLD benchmark results for multi core simulation using four kernels, with varying percentage of remote events.}
\end{figure}

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/phold_priority.eps}
	\caption{PHOLD benchmark results for multi core simulation using four kernels, with varying amount of high-priority events.}
	\label{fig:phold_priority}
\end{figure}

@@

\subsection{Comparison to the abstract simulator}
%reference pSim
The abstract simulator of \textsf{Parallel DEVS} included its own notion of parallelism: transition functions that happen at the same point in simulated time can be trivially parallelized. In this work we refer to this approach as \pSim so as not to confuse with the PDEVS formalism.
This parallelization is trivial, in the sense that there is no need for synchronization protocols.
The parallelization of the transition functions (from now on called the ``inherent parallelism'') is, at least in \textit{dxex}, independent of the use of a synchronization protocol.
That is, both can be used simultaneously.
Following the theoretical analysis published in~\cite{amdahlpdevs}, a comparison between this inherent parallelism and the use of synchronization protocols is warranted.

\paragraph{Model}
We have opted to use the \textit{Queue} model for this comparison, as it allows for many interesting configurations.
%TODO what is meant with this allocation of threads and synchronized kernels?
% 4 Kernels running on a dedicated thread, with per kernel 4 threads allocated to use for PDEVS.
The effect of combining 4 synchronized kernels, with each 4 threads for \pSim, is contrasted with 16 threads for \pSim and 16 kernels for conservative and optimistic.
The first scenario runs each kernel on a dedicated thread synchronized with the other 3 kernels. Each kernel can use up to 4 threads for \pSim. In the second scenario \pSim can use up to 16 threads while running in a single core kernel. Finally 16 synchronized kernels are used with one thread per kernel.
This allows us to observe which is more efficient in obtaining a speedup.
We simulate a computational load by an optional call to sleep in the transition functions set at 5 milliseconds.
The model is configured with depth 4 and width 300 if the transition function has no load, and width 50 if the computational load is active.
In our configuration, an imminent model always generates output, resulting in the receiving model becoming imminent.
The probability that an internal transition in a model generates output and is connected to a receiving model here is $1$. The probability that a model becomes immediately imminent depends on whether fixed or random timeadvance is used. In the case of fixed timeadvance, a model will always become immediately imminent. When random timeadvance is used, this probability is $\frac{1}{n}$, with $n$ denoting the total amount of atomic models. These probabilities correspond with the $q$ parameter in the theoretical analysis ~\cite{amdahlpdevs}. 
The probability that a model is imminent (i.e. is scheduled to execute an internal transition) at any given point in time in this model is equal to $1$ when a fixed timeadvance is used and $\frac{1}{n}$ when a random time advance is used. These probabilities correspond with the $p$ parameter in the theoretical analysis.

The benchmarks are run sufficiently long enough to guarantee that the frequency of internal and external transitions is equal within a benchmark, regardless of the randomness of the timeadvance. In our model each model will be executing internal and external transitions, creating an ideal use case to evaluate the speedup inherent parallelism can obtain.
The key difference is that although the event frequency is the same, their concurrency is not. 
Models are equally distributed over the kernels and between threads in \pSim.
The coefficient of variation of our results is less than $1\%$.
The communication overhead is hard to estimate, but given our coefficient of variation, we can expect this overhead to be constant.

@@

\paragraph{Concurrent events}
First we create a model where all transitions happen simultaneously, with a significant computational load in the transition functions.
In Figure~\ref{fig:pdevs_plot_fixed_sleep}, we observe that exploiting only the inherent parallelism, results in a speedup of about 10.  
Simulation with conservative and optimistic synchronization is run first with 16 kernels, then with 4 kernels having 4 \pSim threads each. 
Both synchronization protocols achieve a speedup similar to only exploiting the inherent parallelism.
Combining the protocols with the inherent parallelism gives near identical results.
In this scenario there is no real advantage between the different parallel configurations.

@@

\paragraph{Computational load}
Finally, we remove the computation load from the transition function, in combination with many concurrent events.
Figure~\ref{fig:pdevs_plot_no_sleep} shows that in single core simulation, the overhead of thread management and shared memory communication is crippling for performance.
Even though many events occur concurrently, the actual computation being done in the transition function does not outweigh the overhead, resulting in much slower simulation than single core execution of transition functions.
When combined with synchronization protocols, performance is still reasonable, though the use of the inherent parallelism is not beneficial either way.
Both synchronization protocols achieve a reasonable speedup without \pSim, as the overhead of thread management (creation and destruction) is avoided.
These results are similar to~\cite{Himmelspach}.

@@

Conservative simulation always uses more memory than single core simulation, as is to be expected.
Additional memory is required for the multiple threads, but also to store all events that are processed simultaneously.

@@

Second, the allocation of models is nontrivial and has a significant impact on performance.
While our multi core speedup for the Queue model, for example, was rather high, this is mostly due to characteristics of the model: the dependency graph does not contain any cycles.
When cycles were introduced, as in the Interconnect model, performance became disastrous.

### 2016-12-03 publications/Journal/6-allocation.tex
Minor corrections.
* turned the pSim into a macro (\pSim) in case we might want to change the name. * replaced a few occurrences of PDEVS with \pSim * changed mentions of "sequential" and "parallel" to single core and multi core where applicable. - - -> might need review to check if I didn't miss anything

The modeller can specify which kernel a model should be allocated to, should such manual intervention be required.
This is handled by the default model allocator.
If no preference is given, a simple striping scheme is used but this is often insufficient.
By overriding the default allocator, a modeller tunes the allocation scheme for a specific model, maximizing parallel speedup.
This interface can be linked to graph algorithms for automatic allocation scheme generation, for example to avoid cycles in the dependency graph.

@@

Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after direct connection.
The PHOLDTree model allows us to investigate parallel speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.

@@

PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain, as was the case in the Queue model.
This topology is interesting since it highlights the effects of allocation.
First, we evaluate the model in sequential simulation to provide a baseline for parallel simulation.

\subsubsection{Sequential Simulation}
Since \textit{adevs} does not use direct connection, we expect a noticeable performance difference between \textit{dxex} and \textit{adevs}.
This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout ($n$) determines the performance penalty \textit{adevs} suffers compared to \textit{dxex}.
Profiling indeed indicates that an increase in width per subtree ($n$) leads to higher overheads in \textit{adevs} due to the lack of direct connection.
\textit{Dxex} uses direct connection, making it independent of fanout.
Performance of \textit{dxex} is, in this model, only dependent on the number of models.
Slight deviations can still be seen, though, caused by the initialization overhead of direct connection.
Both \textit{adevs} and \textit{dxex} scale linearly in the number of atomic models.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/pholdtree_sequential_dn.eps}
    \caption{Effect of hierarchy in sequential simulation.}
    \label{fig:PHOLDtree_seq_dn_benchmark}
\end{figure}

\subsubsection{Parallel Simulation}
Next, we run the model using two different model allocation schemes: breadth-first and depth-first.
But first, we explain what we mean by both allocation schemes.

@@

With depth-first allocation, we traverse the tree in a depth-first way, allocating subsequently visited atomic models to the same node.
This means, intuitively, that subtreees are frequently allocated to the same node.
This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_dfs}.

@@

The breadth first allocation scheme results in a dependency chain with multiple branches, much like in the Queue model.
Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model.
This is not always true though: a single kernel with an unbalanced number of atomic models or unequal computational load in transition functions slows down the remainder of the chain.
This effect is also apparent if the thread a kernel runs on is not fairly scheduled.
With conservative synchronization this leads to excessive polling of the EOT of the other kernels.
With optimistic synchronization this leads to a cascade of rollbacks, since dependent kernels will simulate ahead of the slower kernel.

@@

The most prominent aspect of these results is the low performance for conservative depth-first allocation for two kernels.
This is mostly caused by the difference between a sequential simulation and a parallel simulation: suddenly we need to take into account synchronization and passing around of lookahead values.
And since the number of kernels is low, the overhead dominates any further speedup.
Optimistic is less sensitive to the number of models per kernel as it does not need to poll each model for a lookahead, this explains the lower runtime penalty observed for optimistic.

>>>>

The modeller can specify which kernel a model should be allocated to, should such manual intervention be required.
This is handled by the default model allocator.
If no preference is given, a simple striping scheme is used but this is often insufficient.
By overriding the default allocator, a modeller tunes the allocation scheme for a specific model, maximizing multi core speedup.
This interface can be linked to graph algorithms for automatic allocation scheme generation, for example to avoid cycles in the dependency graph.

@@

Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after direct connection.
The PHOLDTree model allows us to investigate multi core speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.

@@

PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain, as was the case in the Queue model.
This topology is interesting since it highlights the effects of allocation.
First, we evaluate the model in single core simulation to provide a baseline for parallel simulation.

\subsubsection{Single Core Simulation}
Since \textit{adevs} does not use direct connection, we expect a noticeable performance difference between \textit{dxex} and \textit{adevs}.
This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout ($n$) determines the performance penalty \textit{adevs} suffers compared to \textit{dxex}.
Profiling indeed indicates that an increase in width per subtree ($n$) leads to higher overheads in \textit{adevs} due to the lack of direct connection.
\textit{Dxex} uses direct connection, making it independent of fanout.
Performance of \textit{dxex} is, in this model, only dependent on the number of models.
Slight deviations can still be seen, though, caused by the initialization overhead of direct connection.
Both \textit{adevs} and \textit{dxex} scale linearly in the number of atomic models.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/pholdtree_sequential_dn.eps}
    \caption{Effect of hierarchy in single core simulation.}
    \label{fig:PHOLDtree_seq_dn_benchmark}
\end{figure}

\subsubsection{Multi Core Simulation}
Next, we run the model using two different model allocation schemes: breadth-first and depth-first.
But first, we explain what we mean by both allocation schemes.

@@

With depth-first allocation, we traverse the tree in a depth-first way, allocating subsequently visited atomic models to the same node.
This means, intuitively, that subtrees are frequently allocated to the same node.
This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_dfs}.

@@

The breadth first allocation scheme results in a dependency chain with multiple branches, much like in the Queue model.
Such a linear dependency chain can result in a multi core speedup as we demonstrated with the Queue model.
This is not always true though: a single kernel with an unbalanced number of atomic models or unequal computational load in transition functions slows down the remainder of the chain.
This effect is also apparent if the thread a kernel runs on is not fairly scheduled.
With conservative synchronization this leads to excessive polling of the EOT of the other kernels.
With optimistic synchronization this leads to a cascade of rollbacks, since dependent kernels will simulate ahead of the slower kernel.

@@

The most prominent aspect of these results is the low performance for conservative depth-first allocation for two kernels.
This is mostly caused by the difference between a single core simulation and a multi core simulation: suddenly we need to take into account synchronization and passing around of lookahead values.
And since the number of kernels is low, the overhead dominates any further speedup.
Optimistic is less sensitive to the number of models per kernel as it does not need to poll each model for a lookahead, this explains the lower runtime penalty observed for optimistic.
